---
title: "Introduction To R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
```

## R Objects: Overview

According to Chambers (2008), "everything in R is an object" and "nearly everything that happends in R is a function call". But what does this mean for us from a practical standpoint? 

From a high-level view:

- Objects are data structures that can be manipulated
- Every object and its elements can be referred to by either a variable or symbol
- Objects are pointers to R data types represented by structures defined in underlying C code

### Creating R Objects

To create an R object, the assignment operator `<-` is used to assign an R object to a variable. The variable can then be used to reference that R object. 

```{r first-R-object, exercise = TRUE}
# Create a variable called x and assign it the value "Hello World!"
x <- "Hello World!"
x

# Try creating a variable called 'y' and assign it the value 10

# Change the value of x to pi

```

### Data Types

Every R object can be broken down into a set of 'atomic' R objects, referred to as data types. The three primary data types that we will be working with are `numeric`, `character`, and `logical`.

- The `numeric` data type can be further classified as either `integer` or `double`
- The `logical` data type has two distinct values, `TRUE` or `FALSE`

To determine the type of an R object, use the `typeof()` function.

```{r data-types, exercise = TRUE} 
# numeric, L is used to indicate that this is an integer
a_double <- 6.2
a_int <- 10L

# character
comment <- "some text here"

# logical (TRUE or FALSE)
is_it_true <- FALSE

# check type of variables
typeof(a_double)
typeof(a_int)
typeof(comment)
typeof(is_it_true)
```

### Data Structures

A data structure is used to represent the relationship among a set of data elements. Practical implementations of a data structure also provide a set of operations to manipulate the elements in the data structure, including the ability to:

- Access elements
- Add elements
- Modify elements
- Delete elements

R provides the following set of data structures in the base installation:

#### Vector Objects

Vector objects can only contain data elements of the **same data type**. Vector objects include:

  - Scalar (Vector with length 1)
  - Vector
  - Matrix
  - Array
    
#### List Objects

List objects can be composed of data elements of different data types. A special type of list we will use a lot is the data frame. 

In the next sections, we explore each of these data structures.

## R Operators

Before looking at specific R objects, below is a review of the common operators available in R.

### Arithmetic Operators

- +, -, *, /
- Modulus: %%
- Integer Division: %/%
- Exponentiation: ^, **

```{r simple-operators, exercise = TRUE}
# Create two numeric variables (x, y) and assign them values
x <- 3
y <- 7

# Test common arithmetic operators
x + y
x / y
x^2
y %% x
y %/% x
```

### Logical Operators

- <, >, <=, >=
- == (equal)
- != (not equal)
- || (OR), 
- \| (OR, vectorized, element-wise)
- && (AND), 
- & (AND, vectorized, element-wise)

```{r logical-operators, exercise=TRUE}
# Create two numeric variables, x and y, assign them values
x <- 3
y <- 7

# Create two Boolean variables, t and f, assign them values
t <- TRUE
f <- FALSE

# Compare numeric values using logic statements
y > x
y == x

# Compare &&, ||
t && f
t || f
```

### Compare && and || to & and |

- && and || return scalar value using only 1st element of each object being compared
- & and | return logical vector

```{r logical-vector, exercise = TRUE}
a <- c(TRUE, FALSE, TRUE)
b <- c(TRUE, TRUE, FALSE)
a && b
a & b
```

## R Objects: Vectors

### Overview

Vectors are used to store a contiguous set of data elements with the same data type. We already created a few vectors in the previous section (i.e. `a_int <- 6.21`) that are composed of a single element. 

### Creating Vectors

There are multiple ways to create a vector in R, but one of the most common ways is to use the `c()` function to combine a set of data elements into vector object. After a vector is created, the number of elements in the vector can be found by using the `length()` function.

```{r vectors-basic, exercise = TRUE}
# create a numeric vector with 5 values
a_vec <- c(7, 18, 22.2, -4, 0)
a_vec

# length of vector
length(a_vec) 

# vector type
typeof(a_vec) 
```

### Indexing Vectors

To access a single element or multiple elements in a vector, use the `[]` operator. This operator can be used to access elements of a vector through two different methods:

- A numeric vector that contains the indices (positions) of the elements to access
- A logical vector (with length equal to the vector we are accessing) where `TRUE` is used to indicate which elements we want to access

**Note that indices start at one in R**.

```{r vector-indexing, exercise = TRUE}
a_vec <- c(7, 18, 22.2, -4, 0)

# access 3rd element in the vector
a_vec[3]

# access elements 2, 3, 4 with numeric index vector
a_vec[c(2, 3, 4)] 

# access elements 1 and 3 by with logical vector
a_vec[c(TRUE, FALSE, TRUE, FALSE, FALSE)]
```

### Modifying Vectors

With the ability to index into a vector, we can easily change the value for existing elements and add new values to the vector as well. To change the value of exisiting elements, use indexing to access these values and the `<-` operator to assign new values to those elements.

```{r vectors-modify, exercise = TRUE}
a_vec <- c(7, 18, 22.2, -4, 0)
a_vec

a_vec[c(2, 4)] <- c(5, 15)
a_vec
```

To add elements to the end (or middle/beginning) of a vector, `c()` or `append()` can be used.

```{r vectors-modify2, exercise = TRUE}
a_vec <- c(7, 18, 22.2, -4, 0)

more_num <- c(10, 8, 12)

# add more_num to end of a_vec
c(a_vec, more_num)

# add more_num after 2nd element in a_vec
append(a_vec, more_num, after = 2)
```

### Vectorization

A number of operations in R are 'vectorized' to operate on a set of R objects simultaneously. Vectorization not only leads to cleaner and more succinct code, but often a more efficient implementation of the operations as well. 

As an example, a number of simple operations on vectors can be completed in single line of code.

```{r vectorization, exercise = TRUE}
a_vec <- c(7, 18, 22.2, -4, 0)

# add 10 to each element in a_vec
a_vec + 10 

# square each element in a a_vec 
a_vec^2 # square each element

# test whether each element in a_vec is < 0
a_vec < 0

# access all elements with value < 0
a_vec[a_vec < 0] 
```

### Coercion

What happens when we try to use different data types in the same vector?

R will not complain. Rather, all data elements are coerced to a common data type with the following coercion proctocols.

- integer + double -> double
- (integer / double / logical) + character -> character
- logical + integer -> integer
- logical + double -> double

Below is an example of automatic coercion from double to character. 

In addition, we also exemplify another method to define a vector by using `vector()`. This function is useful when we know the length of a vector, but the value for all elements are not known when it is created.

```{r vector-coercion, exercise=TRUE}
# create empty character vector with length 3
a_charvec <- vector(mode = "character", length = 3L)
a_charvec[c(1, 3)] <- c("los angeles", 10)
a_charvec

# check if each element of a_charvec != ""
a_charvec != "" 

# index with Boolean (T/F) vector
a_charvec[a_charvec != ""]
```

## R Objects: Matrices

### Overview

The matrix object is a natural extension of the vector object to two dimensions. One can think of a matrix as being composed of a set of row vectors or a set of column vectors.

From a data analysis perspective, matrices can be viewed as a way to store structured data, where the rows may represent observations and the columns may represent measured features collected for those observations.

### Creating Matrices

To create a matrix in R, use `matrix()`. The three primary values we need to pass to the matrix function are:

- `data`: The set of data elements
- `nrow`: The number of rows in the matrix
- `ncol`: The number of columns in the matrix

By default, a matrix is filled column-by-column using the data elements passed in `data`. Setting `byrow= TRUE` will fill in the matrix row-by-row. In the example below, we can think of the matrix `mat` as being a set of four row vectors or a set of 3 column vectors 

```{r, echo = FALSE}
set.seed(23594)
```

```{r matrices-create, exercise = TRUE}
# Create matrix with 4 rows, and 3 columns
mat <- matrix(data = rnorm(12), nrow = 4, ncol = 3)
mat

# get dimension of the matrix
dim(mat)

# get number of rows and number of columns separately
nrow(mat)
ncol(mat)
```

### Indexing Matrices

Indexing a matrix extends the bracket notation used for vectors by providing the indices we want for each dimension of the matrix `[,]`. The first dimension refers to the row incides and the second dimension refers to the column indices.

The following examples show how to index a matrix using both numeric indices and logical vectors. In addition, one example shows the use of a negation operator `-` to specify a set of incides to remove.

```{r matrices-indexing, exercise=TRUE}
# Create matrix with 4 rows, and 3 columns
mat <- matrix(data = rnorm(12), nrow = 4, ncol = 3)
mat

# element in 4th row and 2nd column
mat[4, 2]

# all elements in 2nd row
# leave blank to get all indices along a dimension
mat[2, ]

# negation operator to remove indices
# returns same as above
mat[-c(1, 3, 4), ]

# select all elements in 1st row and 2nd row with elements < 0.5
mat[1:2, mat[1, ] < 0.5]
```

### Modifying Matrices

As with vectors, modifying elements of a matrix can be done using the `<-` operator. Index the elements of the matrix to be modified and assign an object with the same number of elements as those indexed (and preferably the same dimensions).

```{r matrices-modified, exercise=TRUE}
# Create matrix with 4 rows, and 3 columns
mat <- matrix(data = rnorm(12), nrow = 4, ncol = 3)
mat

mat[1:2, 1:3] <- matrix(1, nrow = 2, ncol = 3)
mat
```

Rows and columns can be added to a matrix by using `rbind()` and `cbind()`.

```{r matrices-modified2, exercise=TRUE}
mat <- matrix(data = rnorm(12), nrow = 4, ncol = 3)
mat

cbind(c(1:4), mat)

rbind(mat, rep(100, 3))
```

## R Objects: Lists

### Overview

Lists are 'generic' vectors that can be composed of different R object types. The structure of lists is very similar to other list or dictionary-type objects in other programming languages (i.e. Python). Lists can be very useful when dealing data that is hierarchical (i.e. from ancestry trees to JSON data). 

### Creating Lists

To create a list object in R, use `list()` and pass a set of `values` or a set of `name = value` pairs. 

```{r lists-create, exercise = TRUE}
# create a list with 3 elements
# - a_vec = vector with values 1, 2, ..., 10
# - random_letters = random sample of 5 letters
# - tell_me_not_true = logical scalar

a_lst <- list(
    a_vec = 1:10,
    random_letters = sample(letters, 5),
    tell_me_its_not_true = FALSE
)
a_lst
```

### Indexing Lists

Accessing specific elements of a list differs from the other objects we have looked at so far. There are three ways that we can access a specific element of a list:

1) `[[{element_name}]]`
2) `[[{element_index}]]`
3) `$` operator

```{r lists_index, exercise = TRUE}
a_lst <- list(
    a_vec = 1:10,
    random_letters = sample(letters, 5),
    tell_me_its_not_true = FALSE
)
a_lst

a_lst[['a_vec']] # using [[{element name}]]

a_lst[[1]] # using [[{element index}]]

a_lst$a_vec # using $
```   

Index operators can be chained together to access elements that are nested within a list.

```{r lists_index2, exercise = TRUE}
a_lst <- list(
    a_vec = 1:10,
    random_letters = sample(letters, 5),
    tell_me_its_not_true = FALSE
)
a_lst

a_lst$a_vec[5] # 5th element of a_vec

a_lst[['random_letters']][2:4] # 1st element of another_lst

a_lst[c('a_vec', 'tell_me_its_not_true')] # access sublist
```

## R Objects: Data Frames

### Overview

A data frame is a special type of list object composed of equal-length vectors. In addition to being able to store various vector types (unlike a matrix), data frames combine the indexing found in both list and matrix objects.The additional functionalities available in data frames make them one of the most popular objects for data analysis and statistical modelling in R.

As with matrices, we can think of the rows of data frame as corresponding to individual observations and the columns as features that were measured for each observation. 

### Creating Data Frames

A data frame can be created by passing a set of `name = vector()` pairs to the `data.frame()` function. The data frame generated below has three columns: `state`, `city`, and `age`. Note the use of  `rnorm()` to generate random draws from a normal distribution with specified mean and standard deviation.

```{r df-create, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 5),
  stringsAsFactors = F
)
dat
```

### Data Frame: Structure

Similar to matrices, the dimensions of the matrix can be found be using `dim()`, `nrow()`, and `ncol()`. 

If we check the data type of the data frame using `typeof()`, it is indeed a list object. The `str()` function provides a summary of an object's overall structure. This function can be very useful when dealing with list objects that have a complicated structure. The `class()` function tells use whether an R object has any additional `class`. Classes help R functions understand how to handle different R objects. 

In the example below, the `data.frame` class tells `nrow()` exactly which version of this fucntion to use for a data frame. 

```{r df-str, exercise = TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)

# dimensions of data frame
dim(dat)
nrow(dat)
ncol(dat)

# additional information for data frame
typeof(dat)
str(dat)
class(dat)

```

### Data Frame: Indexing

All the indexing methods learned previously can be applied to data frames. List-like indexing can be used to access a particular column or set of columns. Matrix-like indexing allows us to easily subset a data frame by both rows and columns.

```{r df-index, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)

# Similar indexing functionality of a list
dat$city

# Subset by rows and columns - bracket operator + data.frame class
dat[1:5, c("city", "age")]
```

### Data Frame: Adding Columns

As with lists, adding new columns can be done by using indexing to specify the new column name and the `<-` operator to pass the values for the new column. 

Remember that the length of the vector object passed should be either one if you want all values to be the same for the new column or equal to the number of rows in the data frame in most cases. 

The `cbind()` function can also be used to add columns to the begin or end of a data frame.

```{r df-add, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)

# get number of rows
n <- nrow(dat)

# simulate age and add to data frame
height_sim <- rnorm(n, mean = 50, sd = 2) 
dat[['height']] <- height_sim 

# simluate gender and add to data frame
gen_sim <- sample(c('M', 'F'), n, replace = TRUE) 
dat$gender <- gen_sim

# add country
dat$country <- "US"

dat
```

### Data Frame: Modify Columns

Columns in a data frame can also be modified by using the index and assignment operators.

```{r df-modify, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)

# add 5 to all ages
dat$age <- dat$age + 5

# change 'SF' to 'San Francisco'

# which indices equal 'SF'
is_sf <- dat$city == "SF"
is_sf

# change those indices
dat$city[is_sf] <- "San Francisco"

dat
```

### Data Frame: Remove Columns

Columns in a data frame can be removed by either setting them to `NULL` or using the negation operator and index (or indices) to be removed. `NULL` is another R object and generally refers to objects that are undefined.

```{r df-remove, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)

dat$age <- NULL # using NULL
dat <- dat[-c(2)] # negation operator and index
dat
```

### Data Frame: Add Rows

The easiest way to add rows to a data frame is the `rbind()` function.

```{r df-add-rows, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)

dat_more <- data.frame(
  state = c("CA", "CA"), 
  city = c("LA", "SF"),
  age = c(50, 35),
  stringsAsFactors = F
)
dat <- rbind(dat, dat_more)
dat
```

## Control Structures

### Overview

Control structures allow us to control the flow of code execution. A number of control structures are available in R to conditionally execute code or repeat the execution of a section of code based on a set of conditions. 

- if/else 
- for loops
- while loops
- do while loops
- repeat
- next
- break
- return

You may be wondering how we have made it this far without using such structures to manipulate the data structures we have encountered. 

The reason for this goes back to idea of 'vectorized' code introduced earlier. 'Vectorization' actually hides some of these control structures in low-level C code. However, there may be cases where it may not be clear how to vectorize a set of operations that we want to apply to a set of data. In these cases, it may be essential to use the core control structures to accomplish our task. 

Here, we will introduce two fundamental structures, if/else and for loops.

### if/else

The `if/else` control structure is used when we want to control whether a block of R code is executed based on a logical expression. The general syntax for a if/else statement is:

```{r if-else, eval = FALSE}
if (logical expression) {
  # run this R code
} else {
  # run this other R code
}
```

Within the `()`, we provide a logical expression that will evaluate to a scalar `TRUE` or `FALSE`. If the expression evaluates to `TRUE`, the code within the `{}` will be executed. If the expression is `FALSE`, the code in second set of `{}` will be executed. A simple example is shown below:

```{r if-else-ex, exercise=TRUE}
num1 <- 10L
num2 <- 15L

num1 == num2

if (num1 == num2) {
  cat(num1, "equals", num2)
} else {
  cat(num1, "does not equal", num2)
}
```

### for loops

A `for` loop executes a block of R code a fixed number of times based on an R object that we provide. The general syntax for a for loop is:

```{r for-loop, eval = FALSE}
for (elem in R_object) {
  # some R code
}
```

In plain English, the code above says to repeat the code in the `{}` for each element (`elem`) in the `R_obect`. As an example, we can print out each element in the vector by accessing the element directly or through indexing:

```{r for-loop-ex, exercise=TRUE}
for (elem in 1:10) {
  print(elem)
}

some_letters <- sample(letters, 10)

for (i in 1:length(some_letters)) {
  print(some_letters[i])
}
```

Notice that we are able to access the current element within the `{}` as we iterate through element of the vector.

### if/else and for loops

Control structures can be combined to iterate through an R object (i.e the column of a data frame) and conditionally execute code based on the value of each element of that R object.

Consider the example in the previous section where we conditionally changed the value of "SF" to "San Francisco". We can accomplish this task using these two control structures.

```{r ifelse-for-ex, exercise=TRUE}
dat <- data.frame(
  state = rep("CA", 8), 
  city = c(rep("LA", 4), rep("SF", 4)),
  age = rnorm(8, mean = 40, sd = 1),
  stringsAsFactors = F
)
dat

n <- nrow(dat)

# using a for loop with indexing
for (i in 1:nrow(dat)) {
  if (dat$city[i] == "SF") {
    dat$city[i] = "San Francisco"
  }
}
dat
```

## Data Wrangling

### Reading Data

Base R and a number of R packages provide the ability to read data into R from a variety of sources. We will start with reading from one of the most common sources used to store data in a tabular format, a `csv` file. 

The `read.csv()` function allows us to read a `csv` file into a `data.frame` from either our local computer or a remote resource (i.e. a file sitting on a server somewhere). Below we exemplify the ability to read a `csv` file from a Github repository. `header = TRUE` specifies that the first row of the file contains the name for each variable (column) in the data set.  

```{r read-data, exercise = TRUE}
file_path <- "https://raw.githubusercontent.com/gmweaver/USCRtutorials/master/inst/tutorials/IntroToR/temp.csv"

temperature <- read.csv(file = file_path, header = TRUE)
temperature

typeof(temperature)
class(temperature)
```

Let's review the structure of the `temperature` data frame.

```{r prepare-temp-data}
file_path <- "https://raw.githubusercontent.com/gmweaver/USCRtutorials/master/inst/tutorials/IntroToR/temp.csv"
temperature <- read.csv(file = file_path, header = TRUE)
```

```{temp-str, exercise=TRUE, exercise.setup="prepare-temp-data"}
str(temperature)
rownames(temperature)
colnames(temperature)
```

### Data Exploration

#### Summarizing Data

Before we start formally analyzing data, we should familiarize ourself with the variables in the data set. The easiest way to start doing this is to compute basic summary statistics for the variables that we are interested in and generate plots that help us better understand the underlying structure of the data.

The `summary()` function provides a quick way to compute a number summary statistics for a variable or a set of variables.

```{r summary-func, exercise=TRUE, exercise.setup="prepare-temp-data"}
summary(temperature$male)
```

These summary statistics are great, but what do they mean in the context of the `male` variable, a categorical variable?

The `male` variable has an `integer` data type but it is actually a binary indicator variable that is coded `1 = 'Male'` and `0 = 'Female'`. With this knowledge, we can now interpret the values computed in `summary()`.

The mean value is actually telling us the proportion of males that are in the data set. The min and max tell us that there is at least one male and one female in our data set. 

Most of statistics computed in `summary()` can be individually computed with other functions.

```{r mean-func2, exercise=TRUE, exercise.setup="prepare-temp-data"}
mean(temperature$male)
median(temperature$male)
min(temperature$male)
max(temperature$male)
```

That's not what we expected, what does `NA` mean?

`NA` is a logical value used internally by R as a missing value indicator. Looking back at the results of `summary()` function, there is in fact one missing value in the `male` variable. Fortunately, we can remove these values when computing these statistics individually.

```{r mean-func, exercise=TRUE, exercise.setup="prepare-temp-data"}
mean(temperature$male, na.rm = TRUE)
median(temperature$male,na.rm = TRUE)
min(temperature$male, na.rm = TRUE)
max(temperature$male, na.rm = TRUE)
```

For categorical variables it might also be nice to see the actual number of males and females by using the `table()` function.

```{r table-func, exercise=TRUE, exercise.setup="prepare-temp-data"}
table(temperature$male)
```

A similar analysis can be done for continuous variables as well. Below are the summary statistics for the `age` variable

```{r summarize-continuous, exercise=TRUE, exercise.setup="prepare-temp-data"}
summary(temperature$age)
mean(temperature$age, na.rm = TRUE)
sd(temperature$age, na.rm = TRUE)
median(temperature$age, na.rm = TRUE)
min(temperature$age, na.rm = TRUE)
max(temperature$age, na.rm = TRUE)
```

#### Plotting Data

In addition to computing simple summary statistics, plotting data helps identify initial patterns and potentially any inconsistencies in our data.

A simple boxplot provides a nice way to visualize the distribution of the `age` variable. We can also create boxplots that are stratified by `male`.

```{r boxplot, exercise=TRUE, exercise.setup="prepare-temp-data"}
boxplot(temperature$age)
boxplot(temperature$age ~ temperature$male)
```

A histogram provides another nice way to summarize the distribution of a continuous variable and can be easily created with the `hist()` function.

```{r histogram, exercise=TRUE, exercise.setup="prepare-temp-data"}
hist(temperature$age)
```

A scatterplot can be used to capture the relationship between two continuous variables by using the the `plot()` function.

```{r scatterplot, exercise=TRUE, exercise.setup="prepare-temp-data"}
plot(mathreported ~ temp, data = temperature)
```

This is just a glimpse of the visualizations that can be created in R using only what is available in the base package. A number of additional options can be used to customize plots and in the future we will explore the `ggplot2` package, an excellent resource to create visualizations in R.
